Evolution of vertical and oblique transmission under fluctuating selection

### 2017-12-19 ms/PNAS/revision.tex
MWF final minor changes before resubmission

In a recent analysis of evolution under fluctuating selection, Xue and Leibler \cite{xue2016evolutionary} allowed an organism to absorb information about the distribution of possible environments by learning the phenotypes of members of its parental lineage from previous generations. They describe this as ``positive feedback that enhances the probability the probability that the offspring expresses the same phenotype as the parent.'' In this formulation there was ``reinforcement of the parent phenotype'' in an offspring, such as might occur through epigenetic inheritance. Although their analysis was not couched in terms of oblique and vertical transmission, as defined by Cavalli-Sforza and Feldman \cite{cavalli1981cultural}, we have been stimulated by their analysis to develop a model in which oblique transmission, at a rate dependent on the trait frequency in the parental generation, occurs in addition to classical vertical transmission. We then ask how fluctuations in selection interact with the rate of oblique transmission to affect evolutionary dynamics and how the rate of oblique transmission itself might evolve.

@@

\noindent{\it Proof}. Let $a=1+\rho{W-w\over w}$ and $b=1+\rho{w-W\over W}$, and our assumption entails  $a>1$ and $0<b<1$. Following \eqref{11}, fixation in $B$ is unstable if $a^kb^l>1$, and similarly  fixation in $A$ is unstable if $a^lb^k>1$. Therefore both fixations are unstable if
\begin{equation}a^kb^l>1\quad\hbox{\tenrm and}\quad a^lb^k>1,\label{18}\end{equation}
or equivalently if
\begin{equation}k\log a +l\log b>0\quad\hbox{\tenrm and}\quad l\log a+k\log b>0.\label{19}\end{equation}
Now the inequalities of \eqref{19} hold if and only if
\begin{equation}k\,{\log(1/b)\over\log a} < l < k\,{\log a\over\log(1/b)}.\label{20}\end{equation}

@@

 In our case there are two constant equilibria $x^*=0$ and $x^*=1$ corresponding to fixation in $B$ and $A$, respectively. We can characterize the stochastic local stability of these fixations as follows.
 \medskip

@@

  \noindent {\bf Result 7}. {\sl Suppose $E[\log(1+\rho s_t)]<0$. Then $x^*=0$, the fixation of phenotype $B$, is stochastically locally stable. In particular, if $E(s_t)\le 0$, $x^*=0$ is stochastically locally stable.}
\smallskip

 The proofs of Results 6 and 7 are in SI Appendices B and C.

@@

 In the  case of the $AkBl$ cycling environment, we write $n=k+l$, and if $w_A=W$, $w_B=w$ for $k$ generations, and $w_A=w$, $w_B=W$ for $l$ generations, we have
 \begin{equation}
 S_n=S_{k+l} =N(k-l)(W-w).
 \label{58}\end{equation}
 Figure 7 shows an example of how $(k-l)$, $\rho$, and $(W-w)$  in \eqref{57} for $u(x)$, interact to affect fixation probabilities. More examples are illustrated in Supplementary Figure S11.

>>>>

In a recent analysis of evolution under fluctuating selection, Xue and Leibler \cite{xue2016evolutionary} allowed an organism to absorb information about the distribution of possible environments by learning the phenotypes of members of its parental lineage from previous generations. They describe this as ``positive feedback that enhances the probability that the offspring expresses the same phenotype as the parent.'' In this formulation there was ``reinforcement of the parent phenotype'' in an offspring, such as might occur through epigenetic inheritance. Although their analysis was not couched in terms of oblique and vertical transmission, as defined by Cavalli-Sforza and Feldman \cite{cavalli1981cultural}, we have been stimulated by their analysis to develop a model in which oblique transmission, at a rate dependent on the trait frequency in the parental generation, occurs in addition to classical vertical transmission. We then ask how fluctuations in selection interact with the rate of oblique transmission to affect evolutionary dynamics and how the rate of oblique transmission itself might evolve.

@@

\noindent{\it Proof}. Let $a=1+\rho{W-w\over w}$ and $b=1+\rho{w-W\over W}$, and our assumption entails  $a>1$ and $0<b<1$. Following \eqref{11}, fixation in $B$ is unstable if $a^kb^l>1$, and similarly  fixation in $A$ is unstable if $a^lb^k>1$. Therefore both fixations are unstable if
\begin{equation}a^kb^l>1\quad\hbox{\tenrm and}\quad a^lb^k>1,\label{18}\end{equation}
or equivalently if
\begin{equation}k\log a +l\log b>0\quad\hbox{\tenrm and}\quad l\log a+k\log b>0.\label{19}\end{equation}
Now the inequalities of (19) hold if and only if
\begin{equation}k\,{\log(1/b)\over\log a} < l < k\,{\log a\over\log(1/b)}.\label{20}\end{equation}

@@

 In our case there are two constant equilibria $x^*=0$ and $x^*=1$ corresponding to fixation in $B$ and $A$, respectively. We can characterize the stochastic local stability of these fixations with the following results, whose proofs are in SI Appendices B and C.
 \medskip

@@

  \noindent {\bf Result 7}. {\sl Suppose $E[\log(1+\rho s_t)]<0$. Then $x^*=0$, the fixation of phenotype $B$, is stochastically locally stable. In particular, if $E(s_t)\le 0$, $x^*=0$ is stochastically locally stable.}
%\smallskip

@@

 In the  case of the $AkBl$ cycling environment, we write $n=k+l$, and if $w_A=W$, $w_B=w$ for $k$ generations, and $w_A=w$, $w_B=W$ for $l$ generations, we have
 \begin{equation}
 S_n=S_{k+l} =N(k-l)(W-w).
 \label{58}\end{equation}
 Figure 7 shows an example of how $(k-l)$, $\rho$, and $(W-w)$  in \eqref{57} for $u(x)$ interact to affect fixation probabilities. More examples are illustrated in Supplementary Figure S11.
